---
title: "Analysis of Variance"
author: "Harsha Achyuthuni"
date: "July 16 2019"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```

In this post, I would like to look into Anova hypothesis testing. The data-set I am going to use is published in https://smartcities.data.gov.in which is a Government of India project under the  National Data Sharing and Accessibility Policy.   

I want to analyse if the unemployment across Bangalore is similar or are there pockets with high unemployment. The [Unemployment_Rate_Bengaluru.csv](https://smartcities.data.gov.in/catalog/city-profile-bengaluru?filters%5Bfield_catalog_reference%5D=2916949&format=json&offset=0&limit=9&sort%5Bcreated%5D=desc) data set has total employed and unemployed people in Bangalore at a Ward level. For simplicity sake, I want to concentrate on three zones, Bangalore-east, Mahadevpura and Rajeshwari Nagar.  

As I want to test that there is significant difference between unemployment rate in different zones in Bangalore, the null and alternate hypothesis will be as follows:  
$$ H_0: \mu_{Bangalore-east} = \mu_{Mahadevpura} = \mu_{Rajeshwari-Nagar} $$  
$$ H_1: Not\, all\, \mu\, are\, equal $$  

Sample unemployment data set: 
```{r activities4, echo=FALSE}
library(tidyverse)
library(e1071)
library(kableExtra)
unemployment <- read_csv('../data/Unemployment_Rate_Bengaluru_ason_01-03-2011_1.csv')
colnames(unemployment) <- c('cityName', 'zoneName', 'wardName', 'wardNo', 'unemployed.no', 'employed.no', 'total.labour')
unemployment <- unemployment %>% mutate(rate = unemployed.no/total.labour) %>% filter(zoneName %in% c('East', 'Mahadevapura', 'Rajarajeswari Nagar'))
kable(dplyr::sample_n(unemployment, 5),
      caption = 'Sample Data') %>% 
  kable_styling(full_width = F) %>%
  collapse_rows(columns = 1:2, valign = "middle")
```

Visualizing the difference of unemployment in each zone.  

```{r activities5, echo=TRUE}
ggplot(unemployment, aes(x=zoneName, y= rate, group = zoneName)) + 
  geom_boxplot() +
  labs(x='Zone', y='Unemployment %') + 
  theme_minimal()
```

The summary statistics of unemployment across different zones are as follows:  
```{r summary_stats, echo=TRUE}
data.summary <- unemployment %>% 
  rename(group = zoneName) %>% 
  group_by(group) %>% 
  summarise(count = n(), 
            mean = mean(rate),
            sd = sd(rate),
            skewness = skewness(rate),
            kurtosis = kurtosis(rate))
data.summary
```

From the above table I observe that the unemployment rate is between 54% and 57% in all the three zones.  

One of the conditions to perform anova is that the population response variable follows a normal distribution in each group. The distribution of unemployment rate in different zones are:

```{r real_plots, echo=TRUE}
ggplot() + 
  geom_density(data=unemployment, aes(x=rate, group=zoneName, color=zoneName), adjust=2) +
  labs(x = 'Unemployment Rate', y='Density',  title = 'Testing normality among response variables') +
  theme_minimal()+theme(legend.position="bottom")
```

From the above graph, it looks like the three groups are normally distributed. To test for normality, we could use Shapiro-Wilk test.  
The test for each group would have the following hypothesis:  
$H_0$: The unemployment rate in the group is normally distributed  
$H_1$: The unemployment rate in the group is not normally distributed 
```{r shaprio-wilk, echo=TRUE}
for(i in 1:nrow(data.summary)){
  test.dist <- (unemployment %>% dplyr::filter(zoneName == data.summary$group[i]))$rate
  cat('Testing for group ', data.summary$group[i], '\n')
  print(shapiro.test(test.dist))
  norm.plot <- ggplot() + 
    geom_qq(aes(sample = test.dist)) +
    stat_qq_line(aes(sample = test.dist)) +
    ggtitle(paste0("Normal distribution Q-Q plot for group ",data.summary$group[i])) +
    theme_minimal()
  plot(norm.plot)
}
```

As $p > \alpha$, where $\alpha = 0.01$, retaining the Null hypothesis in all the three groups. Therefore, we can assume that unemployment rate is normally distributed among all groups.  

Another condition for anova is that the population variances are assumed to be same. This is an assumption I am willing to take at this point.  

Taking these assumptions, the ideal distributions of the sample are as follows. These distributions can be compared to a [simulation](https://harshaash.shinyapps.io/AnovaSimulation/) that I created where change in F value (and significance of anova) can be visualised by increasing the between variance (increasing the distance between group means)  

```{r normal_plots, echo=TRUE}
plot.normal.groups <- function(data.summary, mean, sd, label, title){
  common.group.sd <- mean(data.summary$sd)
  range <- seq(mean-3*sd, mean+3*sd, by = sd*0.001)
  norm.dist <- data.frame(range = range, dist = dnorm(x = range, mean = mean, sd = sd))
  # Plotting sampling distribution and x_bar value with cutoff
  norm.aov.plot <- ggplot(data = norm.dist, aes(x = range,y = dist))
  for (i in 1:nrow(data.summary)) {
    norm.aov.plot <- norm.aov.plot + 
      stat_function(fun = dnorm, color=colors()[sample(50:100, 1)],  size = 1, 
                    args = list(mean = data.summary$mean[i], sd = common.group.sd))
  }
  norm.aov.plot + labs(x = label, y='Density',  title = title) +
      theme_minimal()+theme(legend.position="bottom")
}
set.seed(9)
mean <- mean(unemployment$rate)
sd <- sd(unemployment$rate)
plot.normal.groups(data.summary, mean, sd, 'Travel time (sec)', 'Assuming normality among response variables')
```

Performing Anova with  
$$ H_0: \mu_{Bangalore-east} = \mu_{Mahadevpura} = \mu_{Rajeshwari-Nagar} $$  
$$ H_1: Not\, all\, \mu\, are\, equal $$  

```{r anova-test-function-creation, echo=TRUE}
# Functions used in anova-test

f.plot <- function(pop.mean=0, alpha = 0.05, f, df1, df2,
                              label = 'F distribution',title = 'Anova test'){
  # Creating a sample F distribution
  range <- seq(qf(0.0001, df1, df2), qf(0.9999, df1, df2), by = (qf(0.9999, df1, df2)-qf(0.0001, df1, df2))*0.001)
  f.dist <- data.frame(range = range, dist = df(x = range, ncp = pop.mean, df1 = df1, df2 = df2)) %>% 
    dplyr::mutate(H0 = if_else(range <= qf(p = 1-alpha, ncp = pop.mean, df1 = df1,df2 = df2),'Retain', 'Reject'))
  # Plotting sampling distribution and F value with cutoff
  plot.test <- ggplot(data = f.dist, aes(x = range,y = dist)) +
    geom_area(aes(fill = H0)) +
    scale_color_manual(drop = TRUE, values = c('Retain' = "#00BFC4", 'Reject' = "#F8766D"), aesthetics = 'fill') +
    geom_vline(xintercept = f, size = 2) +
    geom_text(aes(x = f, label = paste0('F = ', round(f,3)), y = mean(dist)), colour="blue", vjust = 1.2) +
    labs(x = label, y='Density',  title = title) +
    theme_minimal()+theme(legend.position="bottom")
  plot(plot.test)
}
```

```{r anova, echo=TRUE}
anva <- aov(rate ~ zoneName, unemployment)
anova.summary <- summary(anva)
print(anova.summary)
f.plot(f = anova.summary[[1]]$F[1], df1 = anova.summary[[1]]$Df[1], df2 = anova.summary[[1]]$Df[2])
```

Here $p < \alpha$, where $\alpha = 0.05$ Hence rejecting the Null Hypothesis. From this test we can observe that not all groups have same mean. But to find out which groups are similar and which are different, I am conducting a TukeyHSD test.  

```{r tukeyHSD, echo=TRUE}
tukey.test <- TukeyHSD(anva)
print(tukey.test)

# Plot pairwise TukeyHSD comparisons and color by significance level
tukey.df <- as.data.frame(tukey.test$zoneName)
tukey.df$pair = rownames(tukey.df)
ggplot(tukey.df, aes(colour=cut(`p adj`, c(0, 0.01, 0.05, 1), 
                           label=c("p<0.01","p<0.05","Non-Sig")))) +
  geom_hline(yintercept=0, lty="11", colour="grey30") +
  geom_errorbar(aes(pair, ymin=lwr, ymax=upr), width=0.2) +
  geom_point(aes(pair, diff)) +
  labs(x = 'Groups', y='Density',  colour="", title = 'Tukey HSD Test') +
    theme_minimal()+theme(legend.position="bottom")
```

From this test we can see that there is significant difference (with $\alpha = 0.05$ confidence) between Rajeshwari Nagar and Bangalore East.    

Created on 17th June 2019, Achyuthuni Sri Harsha