---
title: "Chi Square test of independence"
author: "Harsha Achyuthuni"
date: "July 24 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```

In this post, I would like to look into Chi Square test of independence. The data set I am going to use is published in https://smartcities.data.gov.in which is a Government of India project under the  National Data Sharing and Accessibility Policy.   

I want to find what are the safest and deadliest ways to travel on Bangalore roads. The [Injuries_and_Fatalities_Bengaluru_from_2016to2018.csv](https://smartcities.data.gov.in/catalog/road-safety-bengaluru?filters%5Bfield_catalog_reference%5D=2914260&format=json&offset=0&limit=9&sort%5Bcreated%5D=desc) data set has total number of injuries and fatalities in Bangalore from 2016 to 2018. I want to take injuries as a dummy for the number of incidents that took place.  

As I want to test that there is significant difference in the fatalities with different types of transport, the null and alternate hypothesis will be as follows:   
H0: The type of transport is independent of the fatalities  
H1: The type of transport is dependent  

Sample data set: 
```{r activities4, echo=FALSE}
library(tidyverse)
library(e1071)
data <- read_csv('F:\\System\\Injuries_and_Fatalities_Bengaluru_from_2016to2018.csv')[1,]
data <- data.frame(instance = rownames(t(data)), count = as.numeric(t(data)))
data <- data %>% dplyr::filter(grepl("201",instance))  # Filtering columns with year in it
data$year <- vapply(strsplit(as.character(data$instance), " - ", fixed = TRUE), "[", "", 1)
data$type <- vapply(strsplit(as.character(data$instance), " - ", fixed = TRUE), "[", "", 2)
data$transport <- vapply(strsplit(as.character(data$instance), " - ", fixed = TRUE), "[", "", 3)
data <- data %>% filter(!is.na(transport))
dplyr::sample_n(data, 5)
```

The contingency table for the year 2017 is   

```{r activities5, echo=TRUE}
contingency_table <- data %>% filter(year == 2017) %>% 
  dplyr::select(type, transport, count) %>% 
  spread(type, count)
library(kableExtra)
kable(contingency_table,
      caption = 'Contingency Table') %>% 
  kable_styling(full_width = F) %>%
  column_spec(1, bold = T) %>%
  collapse_rows(columns = 1:2, valign = "middle") %>% 
  scroll_box()

```

A Mosaic plot for the same is:  
```{r summary_stats1, echo=FALSE}
data <- data %>% filter(year == 2017) %>% dplyr::select(type, transport, count)
data$transport <- vapply(strsplit(as.character(data$transport), " ", fixed = TRUE), "[", "", 1)
```


```{r summary_stats, echo=TRUE}
library(ggmosaic)
ggplot(data = data) +
  geom_mosaic(aes(weight = count, x = product(transport), fill = type), na.rm=TRUE) +
  labs(x = 'Type of transport', y='%',  title = 'What type of transport to use') +
    theme_minimal()+theme(legend.position="bottom")
```

From the above plot I can observe that there is a significant difference in the percentages of fatalities in each transport. To find if this percent is significant, I will conduct a chi-square test of independence.  

```{r real_plots, echo=TRUE}
library(gmodels)
# Converting contingency table to flat tables
# Two vectors to hold values of columns
caseType <- c();  conditionType <- c()

# For each cell, repeat the rowname, colname combo 
# as many times
for(i in 1:nrow(contingency_table)) {
  for(j in 2:ncol(contingency_table)) {
    numRepeats <- contingency_table[i, j]
    
    caseType <- append(caseType, 
                       rep(contingency_table[i,1],
                           numRepeats))
    conditionType <- append(conditionType, 
                        rep(colnames(contingency_table)[j],
                            numRepeats))
  }
}

# Construct the table from the vectors
flatTable <- data.frame(caseType, conditionType)
CrossTable(flatTable$caseType, flatTable$conditionType,
           dnn=c("Transportation Type", "Accident type"),
           expected=TRUE)
```

```{r chi-sq-test-function-creation, include=FALSE}
# Functions used in chi-sq-test

chi.sq.plot <- function(pop.mean=0, alpha = 0.05, chi.sq, df,
                              label = 'Chi Square distribution',title = 'Chi Square goodness of fit test'){
  # Creating a sample chi-sq distribution
  range <- seq(qchisq(0.0001, df), qchisq(0.9999, df), by = (qchisq(0.9999, df)-qchisq(0.0001, df))*0.001)
  chi.sq.dist <- data.frame(range = range, dist = dchisq(x = range, ncp = pop.mean, df = df)) %>% 
    dplyr::mutate(H0 = if_else(range <= qchisq(p = 1-alpha, ncp = pop.mean, df = df,lower.tail = TRUE),'Retain', 'Reject'))
  # Plotting sampling distribution and x_bar value with cutoff
  plot.test <- ggplot(data = chi.sq.dist, aes(x = range,y = dist)) +
    geom_area(aes(fill = H0)) +
    scale_color_manual(drop = TRUE, values = c('Retain' = "#00BFC4", 'Reject' = "#F8766D"), aesthetics = 'fill') +
    geom_vline(xintercept = chi.sq, size = 2) +
    geom_text(aes(x = chi.sq, label = paste0('Chi Sq = ', round(chi.sq,3)), y = mean(dist)), colour="blue", vjust = 1.2) +
    labs(x = label, y='Density',  title = title) +
    theme_minimal()+theme(legend.position="bottom")
  plot(plot.test)
}
```


```{r chi-sq, echo=TRUE}
chi.test <- chisq.test(contingency_table[,2:3], rescale.p = TRUE)
print(chi.test)
chi.sq.plot(chi.sq = chi.test$statistic, df = chi.test$parameter, title = 'Null hypothesis to test independence')
```

As p < a, where a = 0.05, I reject the Null hypothesis. There is a significant difference in the mortality rate with different vehicles. Travelling on two-wheeler is the safest while bicycle is the most dangerous.  
Created on 24th June 2019, Achyuthuni Sri Harsha