---
title: "Google Maps Location data"
author: "Harsha Achyuthuni"
date: "December 11, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(dplyr)
library(lubridate)
library(ggplot2)
library(MASS)
set.seed(0512)
attendance <- read.csv("C:/Users/achyuthuni.harsha/Downloads/Attendance_Detail_29_11.csv")
attendance <- attendance %>% 
  filter(Was.at.Work == 'True') %>% 
  mutate(in.time.dmy.hm = dmy_hm(paste(Attendance.Date, In.Time)),
         out.time.dmy.hm = dmy_hm(paste(Attendance.Date, Last.Out.Time)),
         post.midnight = (out.time.dmy.hm - in.time.dmy.hm < 0))
attendance$Attendance.Date <- dmy(attendance$Attendance.Date)

attendance$out.time.dmy.hm[attendance$post.midnight] <- attendance$out.time.dmy.hm[attendance$post.midnight] + days(1)
attendance <- attendance %>% mutate(hours.worked = out.time.dmy.hm - in.time.dmy.hm)

attendance$policy.in.time <- 'char'
msu.intime <- attendance$Attendance.Date > dmy('4/8/2017') & attendance$Attendance.Date <= dmy('17/12/2017')
walmart.intime <- attendance$Attendance.Date > dmy('17/12/2017') & attendance$Attendance.Date <= dmy('30/4/2018')
walmart.buffer.intime <- attendance$Attendance.Date > dmy('30/4/2018') & attendance$Attendance.Date <= dmy('8/6/2018')
walmart.daylight.savings.intime <- attendance$Attendance.Date >= dmy('11/6/2018')
attendance$policy.in.time[msu.intime] <- '1:00 PM'
attendance$policy.in.time[walmart.buffer.intime] <- '1:30 PM'
attendance$policy.in.time[walmart.intime] <- '1:15 PM'
attendance$policy.in.time[walmart.daylight.savings.intime] <- '12:00 PM'

attendance <- attendance %>% 
  mutate(policy.in.time.dmy.hm = ymd_hm(paste(Attendance.Date, policy.in.time))) %>% 
  mutate(policy.out.time.dmy.hm = policy.in.time.dmy.hm + hours(9)) %>% 
  mutate(diff.in.time = difftime(policy.in.time.dmy.hm, in.time.dmy.hm, units = 'mins'),
         diff.out.time = difftime(out.time.dmy.hm, policy.out.time.dmy.hm, units = 'mins'))
```

## Integrating Google location history

I want to integrate my Google Location history to my analysis. This to find the factors  
1. Traffic( Travel time / distance )  
2. Vehicle used

It's no secret that Google stores a data on all of us. But at least they allow us to access quite a lot of the data they have collected on us. Among this is the Google location history.  

The Google location history can be downloaded from your Google account under https://takeout.google.com/settings/takeout. Make sure you only tick "location history" for download, otherwise it will take super long to get all your Google data.  

The data Google provides you for download is a .json file and can be loaded with the jsonlite package. Loading this file into R might take a few minutes because it can be quite big, depending on how many location points Google had saved about you.  


```{r jsonlite, echo=TRUE}
library(jsonlite)
setwd('C:/Users/achyuthuni.harsha/Desktop/Blog/takeout-20181205T104628Z-001/Takeout/Location History')
system.time(
  x <- fromJSON("Location History.json")
)
```

Converting to data frame  
```{r converting_to_df, echo=TRUE}
locations <- as.data.frame(x$locations)
dplyr::sample_n(locations, 5)
```

The date and time column is in the POSIX milliseconds format, so I converted it to human readable format.  

```{r df_modifications_1, echo=TRUE}
# The Unix epoch is 1970-01-01 00:00:00 UTC*
locations$time <- as.POSIXct(as.numeric(locations$timestampMs)/1000, origin = "1970-01-01")
```

Similarly, longitude and latitude are saved in 10^7 format and were converted to GPS coordinates.
```{r df_modifications_2, echo=TRUE}
locations$lat = locations$latitudeE7 / 1e7
locations$lon = locations$longitudeE7 / 1e7
```

The modified dataFrame will look like this
```{r df_print, echo=FALSE}
locations <- locations %>% dplyr::select(time, lat, lon, accuracy, activity)
rm(x)
dplyr::sample_n(locations, 5)
```

### Data stats

Before I get to actually plotting maps, I want to explore a few basic statistics of the data.  
1. How many data points did Google record over what period of time?  
```{r dp_stats, echo=TRUE}
nrow(locations)
```

2. Fron when till when is this data recorded  
```{r dp_stats2, echo=TRUE}
min(locations$time)
max(locations$time)
```

3. How many days were recorded?
```{r dp_stats3, echo=TRUE}
length(unique(date(locations$time)))
```

### Integrating with attendence data set

Finding how many working days are captured
```{r integration1, echo=TRUE}
working_days <- locations %>% 
  filter((date(time) %in% date(attendance$Attendance.Date)))
length(unique(date(working_days$time)))
```

A total of 235 working days were captured of 266 working days.  
```{r integration2, echo=TRUE, message=FALSE, warning=FALSE}
library(ggmap)
qmplot(lon, lat, data = working_days, zoom = 5)
```

Filtering for data in Bangalore
```{r integration3, echo=TRUE, message=FALSE, warning=FALSE}
# Bangalore North Eastern border is 13.071393, 77.787015 (Hoskote)
# Bangalore South Western border is 12.889629, 77.454102 (Kumbalgodu)
bangalore <- locations %>% filter(lon < 77.787015, lon > 77.454102, lat < 13.071393,lat > 12.889629 )
qmplot(lon, lat, data = bangalore, zoom = 14)
```

Considering only the time when I will be leaving the house  
Clustering the longitudes and latitudes near my house. Then finding the maximum time (before the office intime) when I was in this cluster.
```{r integration4, echo=FALSE, message=FALSE}
in_time <- bangalore %>% 
  filter((date(time) %in% date(attendance$Attendance.Date))) %>%
  filter(lon < 77.738302, lat < 12.978126, lon > 77.7) %>% # Left of MuSigma, lower than Cake Mane Right of Decathlon
  filter(hour(time) %in% c(11, 12, 13)) %>% 
  mutate(date = date(time)) %>% 
  group_by(date) %>% 
  summarise(out.of.house.time = max(time),
            out.of.house.lat = max(lat),
            out.of.house.long = max(lon))
# ReStore: 12.975008, 77.755930
# Ganesh Temple: 12.976637, 77.756992
a <- bangalore %>% 
  filter((date(time) %in% date(attendance$Attendance.Date))) %>%
  filter(lon > 77.755930, lat < 12.976637) %>% # Right of ReStore while below Ganesh Temple
  filter(hour(time) %in% c(10, 11, 12, 13)) %>% 
  mutate(date = date(time)) %>% 
  group_by(date) %>% 
  summarise(out.of.house.time = max(time),
            out.of.house.lat = max(lat),
            out.of.house.long = min(lon))
in_time <- rbind(in_time, a) %>%
  group_by(date) %>%
  summarise(out.of.house.time = max(out.of.house.time),
            out.of.house.lat = min(out.of.house.lat),
            out.of.house.long = min(out.of.house.long))
rm(list = c('a', 'bangalore', 'locations', 'working_days'))
dplyr::sample_n(in_time, 5)
```

```{r integration5, echo=FALSE, message=FALSE}
tz(attendance$in.time.dmy.hm) <- 'Asia/Kolkata'
travel <- base::merge(in_time, attendance, by.x = 'date', by.y = 'Attendance.Date')

travel <- travel %>% mutate(travelling.time = in.time.dmy.hm - out.of.house.time) %>% 
  dplyr::select(date, travelling.time) %>% 
  filer(travelling.time > 0, travelling.time < 3600)
dplyr::sample_n(travel, 5)
```

```{r histogram, echo=TRUE}
ggplot(travel,aes(x = as.numeric(travelling.time))) +
  stat_function(fun = dnorm, color="darkred",  size = 1,
                args = list(mean = mean(as.numeric(travel$travelling.time)),
                                         sd = sd(as.numeric(travel$travelling.time)) )) +
  geom_density(aes(y=..density..), color="darkblue",  size = 1)+
  geom_histogram(aes(y=..density..), bins = 50, fill = "red", alpha = 0.2) +
  labs(x = 'Travel Time (seconds)', y='Density') +
  theme_minimal()
```
