---
title: "Missing value imputation using KNN"
author: "Harsha Achyuthuni"
date: "07/07/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem

Real world data is not always clean. Its often messy and contains unexpected/missing values. In this post I will use a non-parametric algorithm called k-nearest-neighbors (KNN) to replace missing values.   

### Data
The data is technical spec of cars. I have taken this data set from [UCI Machine learning repository](https://archive.ics.uci.edu/ml/datasets/auto+mpg) which in turn took it form StatLib library which is maintained at Carnegie Mellon University. The data set was used in the 1983 American Statistical Association Exposition.   
 
```{r cars, echo=FALSE, message=FALSE, warning=FALSE, paged.print=TRUE}
library(tidyverse)
library(RANN)
library(kableExtra)
cars_info <- read.csv('C:\\Users\\Achyuthuni\\Desktop\\BAI\\Tech Material\\Python\\Python_DS_2days\\data/Auto.csv')
cars_info <- cars_info %>% mutate(horsepower = as.numeric(as.character(horsepower)),
                                  origin = as.factor(origin))
kable(dplyr::sample_n(cars_info, 5),
      caption = 'Sample Data') %>% 
  kable_styling(full_width = F) %>%
  collapse_rows(columns = 1:2, valign = "middle") %>% 
  scroll_box()
```

The data set contains the following columns:  
1. mpg: continuous  (miles per gallon)  
2. cylinders: multi-valued discrete 
3. displacement: continuous (cu. inches)  
4. horsepower: continuous  
5. weight: continuous(lbs.)  
6. acceleration: continuous (sec.)  
7. model year: multi-valued discrete (modulo 100)  
8. origin: multi-valued discrete (1. American, 2. European, 3. Japanese)   
9. car name: string (unique for each instance)  

Now I want to find if this data set contains any abnormal values.
```{r find-na, echo=TRUE, paged.print=TRUE}
summary(cars_info)
```

I find that horsepower contains 5 NA values. I can ignore the data points with horsepower NA, or I could impute the NA values using KNN or other methods. Before imputing, I want to make a strong case that my imputation would be right.  
```{r na-cars1, echo=FALSE, paged.print=TRUE}
missing_cars <- cars_info %>% filter(is.na(horsepower)) %>% dplyr::select(mpg, cylinders, displacement, weight, acceleration, year, origin, name)
kable(missing_cars, caption = 'Cars with missing horsepower') %>% 
  kable_styling(full_width = F) %>%
  scroll_box()
```

The assumption behind using KNN for missing values is that a point value can be approximated by the values of the points that are closest to it, based on other variables.  
Let me take three variables from the above data set, mpg, acceleration and horsepower. Intuitively, these variables seem to be related.  
```{r plot, echo=TRUE, paged.print=TRUE}
ggplot(cars_info, aes(x = mpg, y = acceleration, color = horsepower)) + 
  geom_point(show.legend = TRUE) +
  labs(x = 'Mpg', y='Acceleration',  title = "Auto MPG",
       color = 'Horsepower') + 
  scale_color_gradient(low = "green", high = "red",
                       na.value = "blue", guide = "legend") +
  theme_minimal()+theme(legend.position="bottom")
```

In the above plot, the blue color points are null values. I can infer that cars of similar mpg and acceleration have similar horsepower. For a given missing value, I can look at the mpg of the car, its acceleration, look for its k nearest neighbors and get the cars horsepower.  
I am using *preprocess* function in *caret* package for imputing NA's. The *K* value that I am taking is 20 (~ close to square root of number of variables)
```{r knn-imputation, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
library(caret)
preProcValues <- preProcess(cars_info %>% 
                          dplyr::select(mpg, cylinders, displacement, weight, acceleration, origin, horsepower),
                            method = c("knnImpute"),
                            k = 20,
                            knnSummary = mean)
impute_cars_info <- predict(preProcValues, cars_info,na.action = na.pass)
```

The *impute_cars_info* data set will be normalized. To de-normalize and get the original data back: 
```{r na-cars, echo=TRUE, paged.print=TRUE}
procNames <- data.frame(col = names(preProcValues$mean), mean = preProcValues$mean, sd = preProcValues$std)
for(i in procNames$col){
 impute_cars_info[i] <- impute_cars_info[i]*preProcValues$std[i]+preProcValues$mean[i] 
}
```


The imputed horsepower for the missing data points is:  
```{r cars1, echo=FALSE, paged.print=TRUE}
final <- merge(impute_cars_info, missing_cars, by = c('name', 'year', 'origin','mpg', 'cylinders', 'displacement', 'weight', 'acceleration'))
kable(final, caption = 'Imputed data set') %>% 
  kable_styling(full_width = F) %>%
  scroll_box()
```

The actual hp for the cars are as follows:
```{r actual-cars, echo=FALSE, paged.print=TRUE}
final$actual_hp <- c(84,118,100,81,51)
final$difference <- abs(final$actual_hp -final$horsepower)
kable(final %>% dplyr::select(name, year, horsepower, actual_hp, difference), 
      caption = 'Comparison') %>% 
  kable_styling(full_width = F) %>%
  scroll_box()
```

Out of the 5 cars, I was able to impute horsepower for 2 cars with less than 10hp difference, one car within 15hp and two cars within 30hp difference. To get better results, I should use other imputation techniques. Generally these 5 cars are removed while doing any analysis. In R, you could find the removed data set as *mtcars*.   
  
This post is published at [www.harshaash.website/knn-imputation](http://www.harshaash.website/knn-imputation/). For similar posts, please visit [www.harshaash.website](http://www.harshaash.website/log-of-posts/)