---
title: "Linear Algebra - Vectors"
author: "Harsha Achyuthuni"
date: "26/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(geometry)
```

## Topics covered
1. Basics of Vectors  
2. Subspaces and spans  
3. Linear independence of vectors    
4. Basis, Norms, inner products and their properties  


## Basics of vectors
Consider two vectors a and b
```{r}
a <- c(1,2)
b <- c(1,1)
```

The two vectors can be visualised in a 2D coordinate system as follows:
```{r echo=FALSE}
vecs <- data.frame(vname=c("a","b"), 
                   x0=c(0,0),y0=c(0,0), x1=c(1,1) ,y1=c(2,1), 
                   col=1:2)
p <- plot(NA, xlim=c(0,3), ylim=c(0,3), xlab="X", ylab="Y", lwd=3)
p <- with(vecs, mapply("arrows", x0, y0, x1,y1,col=col,lwd=3))
p <- with(vecs, mapply('text', x=x1-.1, y=y1+.1, labels=c('a', 'b') ))
```

We can perform two types of linear operations on these vectors:  
1. Vector addition  
2. Multiplication of the vector with a scalar  

Vector addition:  
If we add a and b together, the sum would be a vector whose members are the sum of the corresponding members from a and b.  
```{r}
a+b
```

```{r echo=FALSE}
vecs <- data.frame(vname=c("a","b", "a+b"), 
                   x0=c(1,0,0),y0=c(1,0,0), x1=c(2,1,2) ,y1=c(3,1,3), 
                   col=1:3)
p <- plot(NA, xlim=c(0,3), ylim=c(0,3), xlab="X", ylab="Y", lwd=3)
p <- with(vecs, mapply("arrows", x0, y0, x1,y1,col=col,lwd=3))
p <- with(vecs, mapply('text', x=(x1+x0)/2, y=(y1+y0)/2, labels=c('a', 'b', 'a+b') ))
```

Vector multiplication:
If we multiply b by 2, we will get a vector with each of its members multiplied by 2.
```{r}
2*b
```

```{r echo=FALSE}
vecs <- data.frame(vname=c("b","2b"), 
                   x0=c(0,0),y0=c(0,0), x1=c(2,1) ,y1=c(2,1), 
                   col=1:2)
p <- plot(NA, xlim=c(0,3), ylim=c(0,3), xlab="X", ylab="Y", lwd=3)
p <- with(vecs, mapply("arrows", x0, y0, x1,y1,col=col,lwd=3))
p <- with(vecs, mapply('text', x=x1-.1, y=y1+.1, labels=c('2b', 'b') ))
```
  
  
## Subspaces and span
As we can perform only two kinds of linear operations on the vectors, any linear combinations will be of the form:  
$$ S= \alpha a + \beta b $$
Where $\alpha$ and $\beta$ are real numbers and $a$ and $b$ are vectors.  

What about all possible linear combinations of $a$  and $b$?  

```{r echo=FALSE}
# library("animation")
# oopt <- ani.options(interval = 0.2, nmax = 100)
# x <- cumsum(rnorm(n = ani.options("nmax")))
# ani.record(reset = TRUE)
# par(bg = "white", mar = c(4, 4, 0.1, 0.1))
# 
# for(i in c(1,0.5, 0.25,0.75, 0, -0.25, -0.5, -0.75, -1)){
#   for(j in c(1.25,1,0.5, 0.25,0.75,0, -0.25, -0.5, -0.75, -1, -1.25)){
#     lin_comb <- i*a+j*b
#     ia <- i*a
#     jb <- j*b
#     vecs <- data.frame(vname=c("a", "transb", "a+b"), 
#                    x0=c(0,ia[1],0),y0=c(0,ia[2],0), x1=c(ia[1],lin_comb[1],lin_comb[1]) ,y1=c(ia[2],lin_comb[2],lin_comb[2]), 
#                    col=1:3)
#     p <- plot(NA, xlim=c(-2,2), ylim=c(-3,3), xlab="X", ylab="Y", lwd=3)
#     p <- with(vecs, mapply("arrows", x0, y0, x1,y1,col=col,lwd=3))
#     p <- with(vecs, mapply('text', x=(x1+x0)/2, y=(y1+y0)/2, labels=c(paste0(i,'a'),paste0(j,'b'),paste0(i,'a+',j,'b') )))
#     ani.record()
#   }
# }
# saveGIF(ani.replay())
# ani.replay()
# ani.options(oopt)

```

The linear combination of the vector a and b form the entire 2D plane.  

Vector space is the space in which the vector can exist. A 2D vector, like a or b, will have vector space of $R^2$ and a 3D vector like d=[1,2,3] is in the vector space $R^3$.   
Span: The set of all possible linear combinations of vectors is called the span of those set of vectors.  

For the above two vectors a = [1,2] and b=[1,1], the entire 2d space is the span, as we can get every vector in the 2d space as a linear combination of the two vectors.  

To explain the difference between vector space and vector span, consider the two vectors
d=[1,2,3] and e=[1,1,1]
As the two vectors are in 3 dimensions, they have a vector space of 3 or $R^3$.  
Adding d+e I get another vector in 3d.  


```{r}
d <- c(1,2,3)
e <- c(1,1,1)
d+e
```
```{r include=FALSE}
library(plot3D)
f <- d+e

vecs <- data.frame(vname=c("d", "transe", "d+e"), 
                    x0=c(0,d[1],0),y0=c(0,d[2],0),z0=c(0,d[3],0), 
                    x1=c(d[1],f[1],f[1]), y1=c(d[2],f[2],f[2]),z1=c(d[3],f[3],f[3]), 
                    col=1:3)
arrows3D(vecs$x0,vecs$y0, vecs$z0, x1 = vecs$x1, y1=vecs$y1 , z1 = vecs$z1)
```

But what are all the linear combinations for d and e?  

```{r echo=FALSE}
# library("animation")
# oopt <- ani.options(interval = 0.2, nmax = 100)
# x <- cumsum(rnorm(n = ani.options("nmax")))
# ani.record(reset = TRUE)
# par(bg = "white", mar = c(4, 4, 0.1, 0.1))
# 
# for(i in c(1,0.5, 0.25,0.75, 0, -0.25, -0.5, -0.75, -1)){
#   for(j in c(1.25,1,0.5, 0.25,0.75,0, -0.25, -0.5, -0.75, -1, -1.25)){
#     f <- i*d+j*e
#     vecs <- data.frame(vname=c("d", "transe", "d+e"), 
#                     x0=0,y0=0,z0=0, 
#                     x1=f[1], y1=f[2],z1=f[3], 
#                     col=1)
#     arrows3D(vecs$x0,vecs$y0, vecs$z0, x1 = vecs$x1, y1=vecs$y1 , z1 = vecs$z1, 
#               xlim = c(-3,3), ylim = c(-3,3), zlim = c(-3,3))
#     ani.record()
#   }
# }
# saveGIF(ani.replay())
# # ani.replay()
# # ani.options(oopt)
# library(rgl)
# plot3d(replicate(2, 1:3), type = 'n', xlim = c(-3, 3), ylim = c(-3, 3), zlim = c(-3, 3), xlab = '', ylab = '', zlab = '')
# planes3d(c(1,1,0),c(2,1,0),c(3,1,0), col = 'red', alpha = 0.6, top=F)
```

This forms a 2d plane which goes thru the origin. Therefore these vectors span a plane (in $R^2$) although their vector space is ($R^3$).  

The maximum span that any set of vectors can have is equal to their vector space.  

## Linear Independence
Linear independence is when one vector has no relationship with another. In the first example with a=[1,2] and b=[1,1], any vectors in the 2d space can be written as a linear combination of a and b. In the second example with d=[1,2,3] and e=[1,1,1], any vector on the plane can be written as a linear combination of d and e. A vector which is not in the plane, like f = [2,3,3] is linearly independent of d and e, as no $\alpha$ and $\beta$ satisfy $f=\alpha d+\beta e$.  

In a vector space of *n* dimensions (vector space is n), there can be at max n vectors which are linearly independent.   

## Bases, norms and inner products
A basis for $R^n$ space is any linearly independent set of vectors S such that span(S) = n.  
From the above examples, a and b are in the vector space $R^2$ and also have their span as $R^2$. Therefore they form a basis for $R^2$.  
Similarly, the three independent vectors d, e and f are in the vector space $R^3$ and form a basis for $R^3$.   

The standard basis for $R^2$ is [1.0] and [0,1].    

The norm of the vector is the length of the vector.

$$ l_2 \,norm(\bar{v}) = \sqrt{x_1^2 + y_1^2+..} $$

```{r}
norm(a,  type="2")
norm(d,  type="2")
```

The dot product (or inner product) takes two vectors as an input and returns a number as an output. It is defined as $\bar{x}.\bar{y} = \sum{x_i\times y_i}$. It represents the length of the shadow of one vector on the other.  

```{r}
library(geometry)
dot(a,b)
```

In the next post in this series, I will talk about matrices.  

References:
1. [Strang, G. (2016). Introduction to Linear Algebra. Wellesley-Cambridge Press](http://math.mit.edu/~gs/linearalgebra/)

