---
title: "Recommendation systems"
author: "Harsha Achyuthuni"
date: "03/07/2019"
output: 
  html_document:
    keep_md: true
---

Association mining is commonly used to make product recommendations by identifying products that are frequently bought together. It is a common technique used to find associations between many variables. It is often used by grocery stores, e-commerce websites, and anyone with large transaction databases. A most common example that we encounter in our daily lives — Amazon knows what else you want to buy when you order something on their site.  
  
Based on the concept of strong rules, [Rakesh Agrawal, Tomasz Imieliński and Arun Swami](http://citeseer.ist.psu.edu/viewdoc/summary?doi=10.1.1.217.4132) introduced association rules for discovering regularities between products in large-scale transaction data recorded by point-of-sale (POS) systems in supermarkets.   

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
```

## Associate mining in R

[Michael Hahsler](https://michael.hahsler.net/) has authored and maintains two very useful R packages relating to association rule mining: the *arules* package and the *arulesViz* package.  

```{r library, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(arulesViz)
library(arules)
library(kableExtra)
```

### Data and EDA

There’s a [public data](http://www.salemmarafi.com/wp-content/uploads/2014/03/groceries.csv) of buying records in a grocery store. The data looks like below.  

```{r read-data, echo=TRUE}
data <- read.transactions('..\\data\\groceries.csv', format = 'basket', sep=',')
inspect(head(data))
```

I want to find the most frequently bought items.  
```{r eda1, echo=TRUE}
itemFrequencyPlot(data, topN = 20, type = "absolute")
```

I want to find the items that are bought frequently together  
```{r eda2, echo=TRUE, paged.print=TRUE}
freq.items <- eclat(data, parameter = list(supp = 0.01, maxlen = 15))
inspect(head(freq.items))
```

### Product recommendation rules

There are three parameters controlling the number of rules to be generated viz. Support, Lift and Confidence.   

Support is an indication of how frequently the item set appears in the data set.   
$$Support = \frac{Number\ of\ transactions\ with\ both\ A\ and\ B}{Total\ number\ of\ transactions} = P\left(A \cap B\right)$$

Confidence is an indication of how often the rule has been found to be true.   
$$Confidence = \frac{Number\ of\ transactions\ with\ both\ A\ and\ B}{Total\ number\ of\ transactions\ with\ A} = \frac{P\left(A \cap B\right)}{P\left(A\right)}$$

Lift is the factor by which, the co-occurrence of A and B exceeds the expected probability of A and B co-occurring, had they been independent. So, higher the lift, higher the chance of A and B occurring together.  
$$Lift = \frac{Confidence}{Expected\ Confidence} = \frac{P\left(A \cap B\right)}{P\left(A\right).P\left(B\right)}$$
 
```{r apriori, echo=TRUE}
rules <- apriori(data, parameter = list(support = 0.0015, confidence = 0.9))
rules_optim <- rules[]
inspect(rules_optim)
```

lhs is “left hand side” and rhs is “right hand side”. In the first low, it’s for the result about “if a customer buy liquor and red/blush wine, which is in lhs column, will the customer buy bottled beer, which is in rhs column?”    
 
To find the item pairs in descending order of support, lift and confidence.    
```{r sort1, echo=TRUE}
inspect(sort(rules_optim, by="confidence", decreasing = T))
inspect(sort(rules_optim, by="support", decreasing = T))
inspect(sort(rules_optim, by="lift", decreasing = T))
```
```{r plot, echo=TRUE}
plot(rules_optim[1:7], method = "graph")
```

By changing the support and confidence cutoffs, we can get better recommendations.  