---
title: "Google Maps Location data"
author: "Harsha Achyuthuni"
date: "December 11, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(dplyr)
library(lubridate)
library(ggplot2)
library(MASS)
set.seed(0512)
attendance <- read.csv("C:/Users/achyuthuni.harsha/Downloads/Attendance_Detail_29_11.csv")
attendance <- attendance %>% 
  filter(Was.at.Work == 'True') %>% 
  mutate(in.time.dmy.hm = dmy_hm(paste(Attendance.Date, In.Time)),
         out.time.dmy.hm = dmy_hm(paste(Attendance.Date, Last.Out.Time)),
         post.midnight = (out.time.dmy.hm - in.time.dmy.hm < 0))
attendance$Attendance.Date <- dmy(attendance$Attendance.Date)

attendance$out.time.dmy.hm[attendance$post.midnight] <- attendance$out.time.dmy.hm[attendance$post.midnight] + days(1)
attendance <- attendance %>% mutate(hours.worked = out.time.dmy.hm - in.time.dmy.hm)

attendance$policy.in.time <- 'char'
msu.intime <- attendance$Attendance.Date > dmy('4/8/2017') & attendance$Attendance.Date <= dmy('17/12/2017')
walmart.intime <- attendance$Attendance.Date > dmy('17/12/2017') & attendance$Attendance.Date <= dmy('30/4/2018')
walmart.buffer.intime <- attendance$Attendance.Date > dmy('30/4/2018') & attendance$Attendance.Date <= dmy('8/6/2018')
walmart.daylight.savings.intime <- attendance$Attendance.Date >= dmy('11/6/2018')
attendance$policy.in.time[msu.intime] <- '1:00 PM'
attendance$policy.in.time[walmart.buffer.intime] <- '1:30 PM'
attendance$policy.in.time[walmart.intime] <- '1:15 PM'
attendance$policy.in.time[walmart.daylight.savings.intime] <- '12:00 PM'

attendance <- attendance %>% 
  mutate(policy.in.time.dmy.hm = ymd_hm(paste(Attendance.Date, policy.in.time))) %>% 
  mutate(policy.out.time.dmy.hm = policy.in.time.dmy.hm + hours(9)) %>% 
  mutate(diff.in.time = difftime(policy.in.time.dmy.hm, in.time.dmy.hm, units = 'mins'),
         diff.out.time = difftime(out.time.dmy.hm, policy.out.time.dmy.hm, units = 'mins'))
rm(list = c('msu.intime', 'walmart.buffer.intime', 'walmart.daylight.savings.intime', 'walmart.intime'))
```

## Integrating Google location history

This blog is a continuation of [in-time analysis problem](http://www.harshaash.website/in-time-problem). I want to analyse my entry time at office and understand what factors effect it.  

I want to integrate my Google Location history to my analysis. This to find the factors  
1. Travel time  
2. Mode of transportation  

It's no secret that Google stores a data on all of us. But at least they allow us to access quite a lot of the data they have collected on us. Among this is the Google location history.  

The Google location history can be downloaded from your Google account under https://takeout.google.com/settings/takeout. I have only selected "location history" for download, otherwise it will take super long to get all my Google data.  

The downloaded data is a .json file and can be loaded with the jsonlite package. Loading this file into R might take a few minutes because it can be quite big, depending on how many location points Google had saved.  


```{r jsonlite, echo=TRUE}
library(jsonlite)
system.time(
  x <- fromJSON("Takeout/Location History/Location History.json")
)
```

Converting to data frame  
```{r converting_to_df, echo=TRUE}
locations <- as.data.frame(x$locations)
dplyr::sample_n(locations, 5)
```

The date and time column is in the POSIX milliseconds format, so I converted it to human readable format.  
```{r df_modifications_1, echo=TRUE}
# The Unix epoch is 1970-01-01 00:00:00 UTC*
locations$time <- as.POSIXct(as.numeric(locations$timestampMs)/1000, origin = "1970-01-01")
```

Similarly, longitude and latitude are saved in 10^7 format and were converted to GPS coordinates.  
```{r df_modifications_2, echo=TRUE}
locations$lat = locations$latitudeE7 / 1e7
locations$lon = locations$longitudeE7 / 1e7
```

The modified dataFrame will look like this
```{r df_print, echo=FALSE}
locations <- locations %>% dplyr::select(time, lat, lon, accuracy, activity)
rm(x)
dplyr::sample_n(locations, 5)
```

### Data stats

Before I get to actually plotting maps, I want to explore a few basic statistics of the data.  
1. How many data points did Google record over what period of time?  
```{r dp_stats, echo=TRUE}
nrow(locations)
```

2. From when till when is this data recorded  
```{r dp_stats2, echo=TRUE}
min(locations$time)
max(locations$time)
```

3. How many days were recorded?
```{r dp_stats3, echo=TRUE}
length(unique(date(locations$time)))
```

### Integrating with attendence data set

The data set contains the time when I swiped into office and the time when I swiped out of office. Data from 4th October 2017 to 29th November 2018 is considered. For more details on attendance data set, please [read here](http://www.harshaash.website/univariate-analysis/)

Finding how many working days are captured in Google Maps.
```{r integration1, echo=TRUE}
working_days <- locations %>% 
  filter((date(time) %in% date(attendance$Attendance.Date)))
length(unique(date(working_days$time)))
```

A total of 246 working days were captured of 266 working days.  
```{r integration2, echo=TRUE, message=FALSE, warning=FALSE}
library(ggmap)
qmplot(lon, lat, data = working_days,zoom = 5)
```

Filtering for data for Bangalore
```{r integration3, echo=TRUE, message=FALSE, warning=FALSE}
# Bangalore North Eastern border is 13.071393, 77.787015 (Hoskote)
# Bangalore South Western border is 12.889629, 77.454102 (Kumbalgodu)
bangalore <- locations %>% filter(lon < 77.787015, lon > 77.454102, lat < 13.071393,lat > 12.889629 )
qmplot(lon, lat, data = bangalore, geom = c("point","density2d"))
```

Considering only the time when I will be leaving the house  
out.of.house.time = Clustering the longitudes and latitudes near my house. Then finding the maximum time (before the office in time) when I was in this cluster.  
in.time = office in time taken from attendance data set  

travelling.time = out.of.house.time - in.time  
```{r integration4, echo=FALSE, message=FALSE}
in_time <- bangalore %>% 
  filter((date(time) %in% date(attendance$Attendance.Date))) %>%
  filter(lon < 77.738302, lat < 12.978126, lon > 77.7) %>% # Left of MuSigma, lower than Cake Mane Right of Decathlon
  filter(hour(time) %in% c(11, 12, 13)) %>% 
  mutate(date = date(time)) %>% 
  group_by(date) %>% 
  summarise(out.of.house.time = max(time),
            out.of.house.lat = max(lat),
            out.of.house.long = max(lon))
# ReStore: 12.975008, 77.755930
# Ganesh Temple: 12.976637, 77.756992
a <- bangalore %>% 
  filter((date(time) %in% date(attendance$Attendance.Date))) %>%
  filter(lon > 77.755930, lat < 12.976637) %>% # Right of ReStore while below Ganesh Temple
  filter(hour(time) %in% c(10, 11, 12, 13)) %>% 
  mutate(date = date(time)) %>% 
  group_by(date) %>% 
  summarise(out.of.house.time = max(time),
            out.of.house.lat = max(lat),
            out.of.house.long = min(lon))
in_time <- rbind(in_time, a) %>%
  group_by(date) %>%
  summarise(out.of.house.time = max(out.of.house.time),
            out.of.house.lat = min(out.of.house.lat),
            out.of.house.long = min(out.of.house.long))
rm(list = c('a', 'locations', 'working_days'))
```

```{r integration5, echo=FALSE, message=FALSE}
tz(attendance$in.time.dmy.hm) <- 'Asia/Kolkata'
travel <- base::merge(in_time, attendance, by.x = 'date', by.y = 'Attendance.Date')

travel <- travel %>% mutate(travelling.time = in.time.dmy.hm - out.of.house.time) %>% 
  filter(travelling.time > 0, travelling.time < 3600)
dplyr::sample_n(travel %>% dplyr::select(date, travelling.time), 5)
```

Plotting histogram of travel time  
```{r histogram, echo=TRUE}
ggplot(travel,aes(x = as.numeric(travelling.time))) +
  geom_histogram(aes(y=..density..), bins = 50, fill = "cornflowerblue", alpha = 0.8) +
  labs(x = 'Travel Time (seconds)', y='Density', title = "Travel time") +
  theme_minimal()
```

### Mode of transport

Google also guesses my activity based on distance traveled per time.   
In the Bangalore data frame, the column activity has activities data.  
Removing null values and creating a data frame for activities (in Bangalore)  
Capturing the activity in main_activity  

```{r activities1, echo=TRUE}

activities <- bangalore$activity

list.condition <- sapply(activities, function(x) !is.null(x[[1]]))

activities  <- activities[list.condition]

activities.df <- do.call("rbind", activities)

activities.df <- activities.df %>%
  mutate(main_activity = sapply(activity, function(x) x[[1]][1][[1]][1]),
         time = as.POSIXct(as.numeric(timestampMs)/1000, origin = "1970-01-01")) %>% 
  dplyr::select(time, main_activity)
dplyr::sample_n(activities.df, 5)
```

Plotting the frequency of Main Activities in Bangalore.

```{r activities2, echo=TRUE}
ggplot(activities.df, aes(x = main_activity, group = main_activity, fill = main_activity)) + 
  geom_bar()  + 
  guides(fill = FALSE) +
  theme_minimal() +
  labs(
    x = "",
    y = "Count",
    title = "Main activities in Bangalore",
    caption = "Associated activity for recorded positions in Bangalore. 
    Because Google records activity probabilities for each position, 
    only the activity with highest likelihood were chosen for each position."
  )
```

Filtering for transportation during coming to office only  
1. Filtering for activities during working days only  
2. Merging the travel data set(contains travel time)  
3. Filtering for time between out.of.house.time and in.time (office in time)  
4. On days where more than one means of transport is used, considering the one which was observed maximum number of times or using some logic.  
```{r activities3, echo=TRUE}
transport <- activities.df %>% 
  filter((date(time) %in% date(attendance$Attendance.Date))) %>% 
  mutate(date = date(time))
transport <- merge(transport, travel, by = 'date') %>% 
  filter(time > out.of.house.time, time < in.time.dmy.hm) %>% 
  group_by(date, main_activity) %>% 
  summarise(count = n()) %>% 
  group_by(date) %>%
  filter(main_activity %in% c("ON_BICYCLE", "EXITING_VEHICLE", 'IN_VEHICLE',"ON_FOOT")) %>%
  top_n(1, count) %>% 
  top_n(1, match(main_activity, c("ON_BICYCLE", 'IN_VEHICLE', "EXITING_VEHICLE", "ON_FOOT"))) %>%
  arrange(date)
```

The final data set will look as follows:
```{r activities4, echo=FALSE}
travel <- merge(transport, travel)
dplyr::sample_n(travel %>% dplyr::select(date, main_activity, travelling.time), 5)
```

Plotting the frequencies of the means of transportation.  

```{r activities5, echo=TRUE}
ggplot(transport, aes(x = main_activity, group = main_activity, fill = main_activity)) + 
  geom_bar()  + 
  guides(fill = FALSE) +
  theme_minimal() +
  labs(
    x = "",
    y = "Count",
    title = "Mode of transport"
  )
```

These variables travel time and mode of transportation will be used in another blog post to do analysis on office in-time.  

###Extra
Scatter plot of travelling time with different means of transportation  

```{r scatter_diagram, echo=FALSE}
ggplot(travel, aes(x=date, y= as.numeric(travelling.time), color = main_activity)) + 
  geom_point(show.legend = TRUE) +
  labs(x = 'Time', y='Travelling Time ( seconds )',  title = "Travel time for various transport over time",
       color = 'Mode of transport') + 
  theme(legend.position="bottom")
```

Scatter plot of in time with various means of transport (During days when Google location data is available)
This can be compared with Index plot in [Univariate-analysis blog](http://www.harshaash.website/univariate-analysis/)  

```{r scatter_diagram2, echo=FALSE}
ggplot(travel, aes(x=date, y= as.numeric(diff.in.time), color = main_activity)) + 
  geom_point(show.legend = TRUE) +
  labs(x = 'Time', y='In-time difference (Minutes)',  title = "Difference of actual in-time and policy in-time",
       color = 'Mode of transport') + 
  theme(legend.position="bottom")
```

Scatter plot of in time for travelling time (During days when Google location data is available)
```{r scatter_diagram3, echo=FALSE}
ggplot(travel, aes(x=date, y= as.numeric(diff.in.time), color = as.numeric(travelling.time))) + 
  geom_point(show.legend = TRUE) +
  labs(x = 'Time', y='In-time difference (Minutes)',  title = "Difference of actual in-time and policy in-time",
       color = 'Travelling time (sec)') + 
  theme(legend.position="bottom")
```


Credits: Some part of the code is taken from [Shring's post](https://shiring.github.io/maps/2016/12/30/Standortverlauf_post).
